When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2018-02-14 paper/paper.tex
Nicerify

We therefore believe it would be unlikely to cause too much astonishment if we suggested that a
better subgraph isomorphism algorithm could be dropped in as a black box replacement in graph
databases systems to improve their performance.  This is not our claim. Instead, this paper shows
that better algorithms invalidate the premise underlying the entire filter~/ verify approach.
We will now demonstrate empirically that filter~/ verify is simply a poor workaround for the kinds
of deficiency in VF2 demonstrated towards the end of the previous section.

@@

>>>>

We therefore believe it would be unlikely to cause too much astonishment if we suggested that a
better subgraph isomorphism algorithm could be dropped in as a black box replacement in graph
databases systems to improve their performance.  This is not our claim. Instead, this paper shows
that better algorithms invalidate the premise underlying the entire filter~/ verify approach.  We
will now demonstrate empirically that filter~/ verify is simply a limited workaround for a subset of
the weaknesses in non-constraint-based subgraph isomorphism algorithms that we demonstrated towards
the end of the previous section.

@@

We saw in the previous sections that although permuting input graphs could improve VF2's behaviour
on some instances, doing so does not make its performance come close to that of domain-based
algorithms.  \citeauthor{DBLP:conf/edbt/KatsarouNT17} say they ``hope that our findings will open up
new research directions, striving to find appropriate, per-query, isomorphic rewritings, in
combination with alternate per-query sub-iso algorithms that can yield large improvements''. We
believe it is important instead to emphasise previous research directions that have already solved
most of this problem, and to help ensure that these techniques become more widely known.

\citet{DBLP:conf/edbt/KatsarouNT17}'s second claim is that ``different
algorithms find different queries hard''. To address this, they run many subgraph isomorphism
algorithms and input permutations in parallel; the extensive literature on parallel portfolios in
general \citep{DBLP:journals/ai/GomesS01}, and the approach by \citet{DBLP:conf/sls/BattitiM07} for
subgraph isomorphism in particular, is not noted (and nor are portfolios mentioned when they say
that ``using machine learning models to predict which version \ldots to employ per query is of high
interest'').  The evidence so far in this paper suggests that we should be wary. It is certainly
true that there are some instances that all algorithms find hard, and there are good theoretical
reasons to believe that these instances genuinely are really hard.  Furthermore, algorithm
portfolios are also well-known to be a successful technique, and are beneficial even with modern
subgraph isomorphism algorithms \citep{DBLP:conf/lion/KotthoffMS16}.  However, in
\cref{section:labelled} we saw that there are many instances that VF2 finds hard that should not be
hard, and that are not hard for other algorithms.  If permuting graphs sometimes addresses the
difficulty of some of these instances, then it is likely that they are not genuinely hard instances
at all, and perhaps it would be better simply to start using an algorithm with domain tracking and
domain-based ordering heuristics, rather than assembling a portfolio of weak algorithms in the hopes
that at least one of them will often get lucky.

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  Suddenly three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This backs up our suspicions: the apparent success of
\citet{DBLP:conf/edbt/KatsarouNT17}'s technique is due to the ease of sometimes getting better
results out of a poor algorithm. Claiming huge improvements from an algorithm portfolio consisting
only of variations of poor algorithms does not demonstrate a genuine improvement over the state of
the art.
